LabMate: a prospectus for types for MATLAB

### 2024-04-12 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] dimensions

%%%% Macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\pct}{\texttt{\symbol{37}}}
\newcommand{\dir}{\texttt{\symbol{62}}}
\newcommand{\res}{\texttt{\symbol{60}}}
\newcommand{\lcb}{\texttt{\symbol{123}}}
\newcommand{\rcb}{\texttt{\symbol{125}}}
\newcommand{\lsb}{\texttt{\symbol{91}}}
\newcommand{\bsl}{\texttt{\symbol{92}}}
\newcommand{\rsb}{\texttt{\symbol{93}}}

@@

Now that we have assembled the data, we may solve the equations and ask for the type of the solution.
\begin{verbatim}
x = M \ voltages
%> typeof x
\end{verbatim}
Did you notice the error in the above code? The vector \texttt{voltages} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead. LabMate detects this error and responds:

\begin{verbatim}
x = M \ voltages
%< unsolved constraint 12 =? 1
%> typeof x
%< the type of x is quite a puzzle
\end{verbatim}
%
That is, LabMate says that for \texttt{x} to have a sensible type, we need \texttt{M} and \texttt{voltages} to have the same number of rows. We forgot to transpose \texttt{voltages}! If we make the necessary repair, we should now see something more satisfying.
\begin{verbatim}
x = M \ voltages'
%> typeof x
%< x :: [ 5 x 1 ] double
\end{verbatim}
This way, LabMate can detect matrix size errors at development time, rather than at runtime --- running the original code would indeed crash with the error:
\[\textcolor{red}{\textbf{\texttt{Matrix dimensions must agree}}}\]

@@

A notable special case is when both $R$ and $C$ are the unit type $R = C = \One$, that is, the type with exactly one element.
%
$\List{\One}$ amounts to a mere `tally', counting indistinct things, which we allow to be written as a number in decimal notation.
%
In this case, there is only one type of entries $E(r,c)$ (because both $r$
and $c$ must be the unique element of the unit type), and the only
information available in the lists $\rs : \List{\One}$ and
$\cs : \List{\One}$ are their lengths.
%
Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
$m \times n$ matrices with entries of type $E$, where $m$ and $n$ are
the unique lists over the unit type of length $m$ and $n$
respectively.
%
That is what we render in LabMate directives and responses as
\[
\lsb \;m \;\texttt{x}\; n\; \rsb\; E
\]
as in our \texttt{M :: [ 1 x 5 ] double} in Section~\ref{sec:example}.

@@

Coming back to the example from Section~\ref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
\begin{verbatim}
%> A :: [ i <- 12,  j <- 5 ]
          (if j == 4 then Q T else Q 1)

%> b :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities. Meanwhile all the entries of \texttt{b} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$. LabMate reports that the type of \texttt{x} should be
\begin{verbatim}
%< x :: [ i <- 5 ]
        (if i == 4
           then Q (ML^2T^(-4)I^(-1))
           else Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first four rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) in the final row.
%
Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{x}, and to leave the type of some of the entries in \texttt{b} as an unknown
\begin{verbatim}
%> b :: [ i <- 5 ]
        (if i == 4
           then ?bTy
           else Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
We can then ask LabMate to solve the \remph{meta variable} \texttt{?bTy}, to which LabMate responds that we must also have \texttt{?aTy} = $\mathsf{Q}\,(\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1})$. This way the types of LabMate can be used to validate and experiment with the dimensional consistency of the quantities involved.

%GravCalc example again: $t_i$ time ($\mathsf{T}$; in seconds); $v_i$ voltage ($\mathsf{ML}^2\mathsf{T}^{-3}\mathsf{I}^{-1}$; in volt)

>>>>

%%%% Macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\pct}{\texttt{\symbol{37}}}
\newcommand{\dir}{\texttt{\symbol{62}}}
\newcommand{\res}{\texttt{\symbol{60}}}
\newcommand{\lcb}{\texttt{\symbol{123}}}
\newcommand{\rcb}{\texttt{\symbol{125}}}
\newcommand{\lsb}{\texttt{\symbol{91}}}
\newcommand{\bsl}{\texttt{\symbol{92}}}
\newcommand{\rsb}{\texttt{\symbol{93}}}
\newcommand{\crt}{\texttt{\symbol{94}}}

@@

Now that we have assembled the data, we may solve the equations and ask for the type of the solution.
\begin{verbatim}
x = M \ voltages
%> typeof x
\end{verbatim}
Did you notice the error in the above code? The vector \texttt{voltages} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead. LabMate will detect this error and respond:

\begin{verbatim}
x = M \ voltages
%< unsolved constraint 12 =? 1
%> typeof x
%< the type of x is quite a puzzle
\end{verbatim}
%
That is, for \texttt{x} to have a sensible type, we need \texttt{M} and \texttt{voltages} to have the same number of rows. We forgot to transpose \texttt{voltages}! If we make the necessary repair, we should now see something more satisfying.
\begin{verbatim}
x = M \ voltages'
%> typeof x
%< x :: [ 5 x 1 ] double
\end{verbatim}
This way, LabMate can detect matrix size errors at development time, rather than at runtime --- running the original code would indeed crash with the error:
\[\textcolor{red}{\textbf{\texttt{Matrix dimensions must agree}}}\]

@@

A notable special case is when both $R$ and $C$ are the unit type $R = C = \One$, that is, the type with exactly one element.
%
$\List{\One}$ amounts to a mere `tally', counting indistinct things, which we allow to be written as a number in decimal notation.
%
In this case, there is only one type of entries $E(r,c)$ (because both $r$
and $c$ must be the unique element of the unit type), and the only
information available in the lists $\rs : \List{\One}$ and
$\cs : \List{\One}$ are their lengths.
%
Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
$m \times n$ matrices with entries of type $E$, where $m$ and $n$ are
the unique lists over the unit type of length $m$ and $n$
respectively.
%
That is what we render in LabMate directives and responses as
\[
\lsb \;m \;\texttt{x}\; n\; \rsb\; E
\]
as in our \texttt{M :: [ 12 x 5 ] double} in Section~\ref{sec:example}.

@@

Coming back to the example from Section~\ref{sec:example} again, we aim to give both \texttt{M} and \texttt{voltages} more precise types, also taking their physical dimensions into account,
\begin{verbatim}
%> M :: [ 12 x i::D<-[{},{},{},{},T] ] Q(i)
%> voltages :: [ 1 x 12 ] Q(V)
\end{verbatim}
  but we had better unpack what we mean.

Firstly, \texttt{M}'s row header type remains $\One$, so our row header list is still abbreviated by a number.

Secondly, we see \texttt{Q(}$d$\texttt{)} as the type of \remph{quantities} with physical dimension $d$, represented in MATLAB as a mere \texttt{double} with standardised units. We shall define a type $\texttt{D}$ of dimensions presently. The notation
\begin{verbatim}
i::D<-[{},{},{},{},T]
\end{verbatim}
will tell LabMate that the column header type is $\texttt{D}$, and that the matrix entry type \texttt{Q(i)} varies in accordance with the header, locally named \texttt{i}, which is drawn from the given list of dimensions.
%
But what type is \texttt{D}? We will have defined, in LabMate directives, rather than MATLAB,
\begin{verbatim}
%{>
Base =
  ['Time, 'Length, 'Mass, 'Current,
   'Temperature, 'Amount, 'Luminosity]
D = Abel(Enum(Base))
%}
\end{verbatim}
where \texttt{Base} is a list of symbolic constants standing for base dimensions, \texttt{Enum} converts any such list to a finite enumeration type, and \texttt{Abel} constructs the free Abelian group on any type of generators, with unit \lcb\rcb, singleton generators \lcb$g$\rcb, multiplication \texttt{*}, division \texttt{/}, and exponentiation by integer constants \crt. We further define
\begin{verbatim}
%{>
T = {'Time}
V = {'Mass}*{'Length}^2/T^3/{'Current}
%}
\end{verbatim}
and \texttt{voltages} becomes a row vector uniformly containing quantities in \texttt{Q(V)}.

With this refinement in place
\begin{verbatim}
x = M \ voltages'
%> typeof x
\end{verbatim}
will yield
\begin{verbatim}
%< x :: [ i::D<-[{},{},{},{},T] ] Q(V/i)
\end{verbatim}

In other words, \texttt{x} is a column vector comprising four quantities in \texttt{Q(V)} and a fifth in \texttt{Q(V/T)}.
LabMate has thus propagated the dimensional variation from the type of \texttt{M} to the type of \texttt{x}, while retaining the uniformity of \texttt{V} in the type of \texttt{voltages}. Checking multiplicability for matrices of quantities \texttt{Q(}$d(i,j)$\texttt{)} and
\texttt{Q(}$d(j,k)$\texttt{)} amounts to checking that $d(i,j)\texttt{*}d(j,k)$ is independent of $j$. In our example, there is no dimensional variation in the `middle' of the inputs, so multiplicability is evident. There is a degree of freedom to multiply headers by an arbitrary dimension and compensate in the entry type, but we avoid ambiguity by sticking to Hart's discipline~\cite{hart}, namely
\begin{enumerate}
\item ensure any list of dimensions starts with the unit, \lcb\rcb;
\item divide by dimensions coming from a row header;
\item multiply by dimensions coming from a column header.
\end{enumerate}
Valid operations which respect Hart's discipline will always have header lists which meet in the middle exactly and yield outputs which retain Hart's discipline~\cite{dimTypes}.
